Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2014-09-23 Article.tex
A few corrections
optimization problem equation and other updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study coalition formation of prosumers that aim at selling energy to the grid. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and minimum production requirement in order to enter the energy market. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level which leads to finding uncorrelated sources of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

There seems to have a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as most of the demand side management concepts have proposed such an architecture \cite{WuHamedHuangBook2011}. Nevertheless as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (such as coalitions) that will help to decrease the communication burden and in the mean time satisfy the multiple requirements of the power grid management. The formation of coalitions inside the smart grid environment could be applied to various types of agents such as self sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in some emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the mean time reduce the communication burden.

@@

As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. In this paper, we consider that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

@@

\subsection{Coalition formation model}\label{subsec:Coalition}
More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value at agent i at instant t (its instantaneous production minus his instantaneous consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected (because we introduced them), the time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both of the energetic mix and the habits of a given prosumer. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i, and by $ \mathcal{P}_{i} $ the corresponding probability distribution.

We now extend these notations to any coalition $ S \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $
\item $ \mathcal{T}_{S} = \{ P_{S}(t_{0}),...,P_{S}(t_{K}) \} $ with $ \mathcal{P}_{S} $ the corresponding probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{S}^{CRCT} $ on the market.
\end{itemize}

Due to the volatile nature of DER productions or user consumptions, the available power of a coalition will oscillates around $ P_{S}^{CRCT} $. It will be the responsibility of the coalition, that is the information system maintaining the coalition, to compensate for deviations. In this paper, we consider then that the smaller these deviations, the better. Based on the $ \mathcal{T}_{S} $ distribution, the probability of S value (at any instant t) being below its contract value $ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] $ appears therefore as a quantity of interest that we would want to keep as low as possible.

In this model, we consider that only the coalitions that have this probability value sufficiently low are acceptable. That is, we denote by $ \phi \in [0,1] $ the threshold imposed to the coalitions over this quantity, and we call it the \textbf{reliability} in the following. The reliability stipulates then that for any coalition S willing to join the market, the probability of S value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall S,\ \forall t,\ Pr[P_{S}(t) \leq P_{S}^{CRCT} ] \leq \phi $ ($ \phi $ being restricted to small values for consistency).

We then consider another constraint denoted by $ P^{MIN} $, as the \textbf{minimum contract value}, which states that only coalitions with higher contract values ($ P_{S}^{CRCT} \geq P^{MIN} $) will be accepted. In table \ref{table1} summarizes the main notations of this paper.

At this point, if a coalition S wishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(S) $). If $ P_{\phi}(S) \geq P^{MIN} $, meaning that it also obeys the minimum contract value rule, then S announces this value on the market : $ P_{S}^{CRCT} = P_{\phi}(S) $. Otherwise, coalition S is not able to enter.
Basically, a coalition S is valid if and only if :
\begin{equation}
\left\{ \begin{array}{lll}
		\forall t,\ Pr[ P_{S}(t) \leq P_{\phi}(S)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(S) \geq P^{MIN}\ \textit{{\scriptsize (min value rule)}}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
We now choose a very simple utility function (eq. \ref{utility}) that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|S| $ term in eq. \ref{utility} indicates that we favorise small coalitions, mainly because they are easier to maintain in terms of communications.

\begin{equation}
 \mathcal{U}_{\phi,\ P^{MIN}}(S) = \mathbf{1}_{\textit{S\ valid}} \dfrac{P_{\phi}(S)}{|S|} 
\label{utility}
\end{equation}

@@

In order to illustrate what is done in the following, lets consider a very simple example with two agents, say i and j, with gaussian value probability distributions $ \mathcal{P}_{i} = \mathcal{N}(\mu_{i}, \sigma_{i}) $ and $ \mathcal{P}_{j} = \mathcal{N}(\mu_{j}, \sigma_{j}) $, such that the joint probability distribution  $ \mathcal{P}_{ij} $ of the coalition $\{ij\}$ is also a gaussian with the following parameters :

@@

with $ \rho_{ij} $ the Pearson correlation coefficient between $ P_{i} $ and $ P_{j} $. We can easily write the reliability condition $ Pr[P_{ij}(t) \leq P_{\phi}(ij) ] \leq \phi $ as :

@@

What this paper investigates in the following is the developement of an algorithm that organizes prosumers such that the synergy term of the standard deviation ( $ \rho_{ij}\sigma_{i}\sigma_{j} $ in the example above) is minimized. In such settings, and through the clique percolation procedure (section \ref{sec:forming}), we will show that coalitions with low relative standard deviations, e.g high utility coalitions, can be computed.

\begin{table}[!t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  $ \mathcal{A}=\{a_{1},...,a_{N}\} $ & Set of agents \\
  $ P_{i}(t) \in \mathbb{R}\ \ (P_{S}(t)) $ & Instantaneous production of agent $i$ at time $t$ \\
  $ \mathcal{T}_{i} = \{P_{i}(t_{0}),...,P_{i}(t_{K})\} $ & Time series of agent i available production values \\
  $ \mathcal{P}_{i}\ \ (\mathcal{P}_{S}) $ & Probability distribution of the production of the agent i \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{S}^{CRCT} \in \mathbb{R}^+ $ & Contract value of coalition S \\
  $ P_{\phi}(S) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(S) $ & Utility function \\
  $ \tau_{i}(S) \in \mathbb{R}$ & Decision rule for mapping overlapping into disjoint coalitions \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of desired coalition \\
  $ CS = \{S_{1},...,S_{N_{COAL}}\} $ & Coalition structure \\
  \hline
\end{tabular}

@@

The model we used in order to simulate timeseries of available production provides some diversity because of the combination between the energetic mix and the climate vectors. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is apparent when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $), where well defined clusters appear in the $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques, if it is too low, the $\epsilon$-graph of $ G_{2} $ will not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally repulsive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(S) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore.

Cliques for small values of $ \epsilon $ appear thus as good seeds for stable coalitions, but we still have to find a way to bring them above the grid requirements as well as increasing the social welfare, both by incorporating more nodes. This is exactly where clique percolation proves useful for our concerns. The main idea of making the seeds grow as long as it increases a fitness function is exactly what we need. The only difference being that our fitness function is simply chosen as the utility function.

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions separated and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that consider nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition S and a node $ i \in S $, we define :

\begin{equation}
\tau_{i}(S) = \dfrac{\mathcal{U}_{\phi,\ P^{MIN}}(S) - \mathcal{U}_{\phi,\ P^{MIN}}(S-\{i\}) }{\mathcal{U}_{\phi,\ P^{MIN}}(S)}
\label{tau}
\end{equation}

If node i belongs to  multiple coalitions, the only coalition retaining node i is the one that maximizes $ \tau_{i} $.

At this point, we have three degrees of freedom : the reliability ($ \phi $), the required power to enter the market ($P^{MIN}$), and the prunning parameter ($\epsilon$). For convenience, we introduce the number of desired coalitions ($ N_{COAL} $) and we fix $\epsilon $ as the smallest value such that the $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

The next section shows how the utility behave within the parameters space and provides the results of our 200 agents test.

@@

As shown in figure \ref{fig2:a}, the parameters $\phi$ and $P^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high. Only small values of $ P^{MIN}$ could then potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and more valid coalitions can arise for a larger spectrum of $ P^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because the coalitions formed are able then to announce higher contract values, yielding higher utilities.

In figure \ref{fig2:b}, we fix the reliability value to a given empirical value ($\phi = 0.08 $). We can observe how the social welfare $\sum_{S \in CS} \mathcal{U}_{\phi,\ P^{MIN}}(S)$ evolves according to $P^{MIN}$ and $ N_{COAL} = |CS| $. In our setting we can observe  a maximum point around $ N_{COAL} = 3 $, followed by a decrease. There is clearly a maximum number of coalitions sustainable for a given set of parameters. After that point we will form more coalitions that would be unable to pass the grid requirements. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs, meaning that the algorithm performance will also decrease.

@@

In figure \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{coal} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm performs better at finding high valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in figure \ref{Fig3}).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Abstract
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In a smart grid environment, we study coalition formation of prosumers that aim at entering the energy market. It is paramount for the grid operation that the energy producers are able to sustain the grid demand in terms of stability and minimum production requirement. We design an algorithm that seeks to form coalitions that will meet both of these requirements:  a minimum energy level for the coalitions and a steady production level which leads to finding uncorrelated sources of energy  to form a coalition. We propose an algorithm that uses graph tools such as correlation graphs or clique percolation to form coalitions that meet such complex constraints. We validate the algorithm against a random procedure and show that, it not only performs better in term of social welfare for the power grid, but also that it is more robust against unforeseen production variations due to changing weather conditions for instance. 
\end{abstract}

@@

There seems to have a clear consensus on the benefit of having bidirectional communication flows between the prosumers and the grid, as demand side management relies on such an architecture \cite{WuHamedHuangBook2011}. Nevertheless as the number of active prosumers is expected to rise, it is safe to assume there is a need for more complex communication patterns (such as coalitions) that will help to decrease the communication burden and satisfy the multiple requirements of the power grid management. The formation of coalitions inside the smart grid environment could be applied to various types of agents such as self sufficient micro-grids \cite{Pahwa}, sizable and adjustable virtual power plants (VPP) \cite{Braun, Ramchurn}, or fleet of electric vehicles that back up the grid in some emergency situations (V2G) \cite{Ramchurn}. These are only a few use cases where the coalition of multiple agents can enhance the grid reliability and in the mean time reduce the communication burden.

@@

As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. We consider here that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (air conditioning is not considered). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

@@

\subsection{Coalition formation model}\label{subsec:Coalition}
More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value at agent i at instant t (its instantaneous production minus his instantaneous consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected, the prosumer's time-series exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energetic mix and the habits. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting time-series for  agent i, and by $ \mathcal{F}(P_{i}) $ the corresponding probability distribution.

We now extend these notations to any coalition $ s \subset \mathcal{A} $ : 
\begin{itemize}
\item $ P_{s}(t) = \sum_{i \in s} P_{i}(t) $
\item $ \mathcal{T}_{s} = \{ P_{s}(t_{0}),...,P_{s}(t_{K}) \} $ with $ \mathcal{F}(P_{s}) $ the corresponding probability distribution
\item A coalition S has the possibility to announce a contract value $ P_{s}^{CRCT} $ on the market.
\end{itemize}

Due to the volatile nature of DER productions or user consumptions, the available power of a coalition will oscillates around $ P_{s}^{CRCT} $. It is of the responsibility of the information system that forms coalitions to select agents in order to provide a production as stable as possible. Based on the $ \mathcal{F}(P_{s}) $ distribution, the probability of s value being below its contract value $ Pr[P_{s} \leq P_{s}^{CRCT} ] $ appears therefore as a quantity of interest that we would want to keep as low as possible.

More formally, the rest of the paper derives a heuristic that aims at solving the following problem (Table \ref{table1} summarizes the main notations):
 \begin{equation}
  \min_{\substack{S \subset CS\\ |S| = N_{COAL} \\ 
     \forall s \in S,\ |s| \neq 0 \\ P_{s}^{CRCT} \geq P^{MIN}}}
  \Bigg\{ \sum_{s \in S} Pr[P_{s} \leq P_{s}^{CRCT}] \Bigg\}
  \label{problem}
\end{equation}

We explain now the two requirements that will help us solve this problem. Let $ \phi \in [0,1] $ be the \textbf{reliability} threshold imposed to the coalitions. It stipulates that for any coalition s willing to join the market, the probability of s value (at any instant t) being below its contract value should at most be $ \phi $. That is, $ \forall s \in 2^{N},\ Pr[P_{s} \leq P_{s}^{CRCT} ] \leq \phi $ ($ \phi $ being restricted to small values for consistency).

We also consider the \textbf{minimum contract value} constraint denoted by $ P^{MIN} $, which states that only coalitions with higher contract values ($ P_{s}^{CRCT} \geq P^{MIN} $) will be accepted.

At this point, if a coalition s wishes to join the market, it has to choose a contract value that fulfill the two conditions. For simplification, we consider that coalitions will always apply the same economically consistent strategy of announcing the highest possible contract value that obeys the reliability rule (a value we denote by $ P_{\phi}(s) $). If $ P_{\phi}(s) \geq P^{MIN} $, meaning that it also obeys the minimum contract value rule, then s announces this value on the market : $ P_{s}^{CRCT} = P_{\phi}(s) $. Otherwise, coalition s is not able to enter.
Basically, a coalition s is valid if and only if :
\begin{equation}
\left\{ \begin{array}{lll}
		Pr[ P_{s} \leq P_{\phi}(s)] \leq \phi\ \textit{{\scriptsize (reliability rule)}} \\
		and\ P_{\phi}(s) \geq P^{MIN}\ \textit{{\scriptsize (min value rule)}}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
We now choose a utility function (eq. \ref{utility}) that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|s| $ term in eq. \ref{utility} indicates that we favor small coalitions, mainly because they are easier to maintain in terms of communications.

\begin{equation}
 \mathcal{U}_{\phi,\ P^{MIN}}(s) = \mathbf{1}_{\textit{s\ valid}} \dfrac{P_{\phi}(s)}{|s|} 
\label{utility}
\end{equation}

@@

In order to illustrate what is done in the following, lets consider a very simple example with two agents, say i and j, with gaussian value probability distributions $ \mathcal{F}(P_{i}) = \mathcal{N}(\mu_{i}, \sigma_{i}) $ and $ \mathcal{F}(P_{j}) = \mathcal{N}(\mu_{j}, \sigma_{j}) $, such that the joint probability distribution  $ \mathcal{F}(P_{ij}) $ of the coalition $\{ij\}$ is also a gaussian with the following parameters :

@@

with $ \rho_{ij} $ the Pearson correlation coefficient between $ P_{i} $ and $ P_{j} $. We can easily write the reliability condition $ Pr[P_{ij} \leq P_{\phi}(ij) ] \leq \phi $ as :

@@

What this paper investigates in the following is the developement of an algorithm that organizes prosumers such that the synergy term of the standard deviation ($ \rho_{ij}\sigma_{i}\sigma_{j} $ in the example above) is minimized. In such settings, and through the clique percolation procedure (section \ref{sec:forming}), we will show that coalitions with low relative standard deviations, e.g high utility coalitions, can be computed.

\begin{table}[!t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  \\
  $ \mathcal{A}=\{a_{1},...,a_{N}\} $ & Set of agents \\
  $ P_{i}(t) \in \mathbb{R}\ \ (P_{s}(t)) $ & Instantaneous production of agent $i$ at time $t$ \\
  $ \mathcal{T}_{i} = \{P_{i}(t_{0}),...,P_{i}(t_{K})\} $ & Time series of agent i available production values \\
  $ \mathcal{F}(P_{i})\ \ (\mathcal{F}(P_{S})) $ & Probability distribution of the production of the agent $i$ \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{s}^{CRCT} \in \mathbb{R}^+ $ & Contract value of coalition s \\
  $ P_{\phi}(s) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(s) $ & Utility function \\
  $ \tau_{i}(s) \in \mathbb{R}$ & Decision rule for mapping overlapping into disjoint coalitions \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of desired coalition \\
  $ CS $ & Set of all coalition structures \\
  \\
  \hline
\end{tabular}

@@

The model we used in order to simulate timeseries of available production provides some diversity because of the combination between the energetic mix (renewable generators combination) and the climate vectors. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is apparent when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $), where well defined clusters appear in the $ \epsilon $-graph for any values of $ \epsilon $.

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques, if it is too low, the $\epsilon$-graph of $ G_{2} $ will not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally repulsive. Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ seems too brutal. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore.

Cliques for small values of $ \epsilon $ appear thus as good seeds for stable coalitions. However we have to find a way to bring the coalitions above the grid requirements as well as increasing the social welfare by incorporating more nodes to the coalitions. Clique percolation algorithms have proven to be a useful heuristic for such complex problems. The main idea behind this algorithm is to make the seeds grow as long as they increase a fitness function based on internal and external degrees of the seeds \cite{Lancichinetti}. The main difference between our case study and that algorithm is that we chose this fitness function as our utility function (cf. eq.\eqref{utility}) that both maximizes the production and minimizes its volatility.

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions disjoint and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that considers nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition s and a node $ i \in s $, we define :

\begin{equation}
\tau_{i}(s) = \dfrac{\mathcal{U}_{\phi,\ P^{MIN}}(s) - \mathcal{U}_{\phi,\ P^{MIN}}(s-\{i\}) }{\mathcal{U}_{\phi,\ P^{MIN}}(s)}
\label{tau}
\end{equation}

If node i belongs to  multiple coalitions, the only coalition including node i is the one that maximizes $ \tau_{i} $.

At this point, we have three degrees of freedom : the reliability ($ \phi $), the required power to enter the market ($P^{MIN}$), and the prunning parameter ($\epsilon$). For convenience, we introduce the number of desired coalitions ($ N_{COAL} $) and we choose $\epsilon $ as the smallest value such that the $ \epsilon $-graph contains at least $ N_{COAL} $ non overlaping cliques. The overall utility depends now on the grid requirements and on the number of desired coalitions.

Next section shows the utility behavior within the parameters space and provides the results of our 200 agents test.

@@

As shown in figure \ref{fig2:a}, the parameters $\phi$ and $P^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high. Only small values of $ P^{MIN}$ could then potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and more valid coalitions can arise for a larger spectrum of $ P^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because the coalitions formed are able to announce higher contract values, yielding higher utilities.

In figure \ref{fig2:b}, we fix the reliability value to a given empirical value ($\phi = 0.08 $). We can observe how the social welfare $\sum_{s \in S} \mathcal{U}_{\phi,\ P^{MIN}}(s)$ evolves according to $P^{MIN}$ and $ N_{COAL} $. In our setting we can observe  a maximum point around $ N_{COAL} = 3 $, followed by a decrease. There is clearly a maximum number of coalitions sustainable for a given set of parameters. After this value we will form more coalitions that would be unable to pass the grid requirements. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs, meaning that the algorithm performance will also decrease.

@@

In figure \ref{Fig3}, we perform a simulation with 200 prosumers, and we fix the reliability $ \phi = 0.1 $ meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We observe the result of our algorithm for a fixed number of coalitions of $N_{COAL} = 10$ and let the algorithm form the coalitions as the grid changes according to $ P^{MIN} $. We compare our algorithm with a random process that splits the agents in $ N_{COAL} $ coalitions. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the \textit{"under-requirement space"}, meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm performs better than the random formation algorithm at finding high valued coalitions. The blue dots allowed to enter the market (non hatched zone) are indeed outnumbering the red dots, especially when the grid requirements are neither too low nor too high ($ P^{MIN} = 5 $ in figure \ref{Fig3}).
